---
title: "2021-10-18 TEP Data Normalization"
output: html_notebook
---

```{r}
# install required CRAN packages
for (pkg in c("BiocManager", "data.table", "httr", "FactoMineR", "dendextend", "googledrive")) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
}

# install required Bioc packages
for (pkg in c("limma", "edgeR")) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    BiocManager::install(pkg, update = FALSE, ask = FALSE)
  }
}
```

## Load data

```{r}
library(data.table)

DATA <- "data"
count_file_path <- "GSE68086_TEP_data_matrix.txt.gz"
meta_file_path <- "tep_gse68086_meta.txt"

rc <- data.frame(fread(file.path(DATA, count_file_path)),
                 row.names = 1, check.names = FALSE)
meta <- read.table(file.path(DATA, meta_file_path),
                   sep = "\t", check.names = FALSE,
                   row.names=1, stringsAsFactors = FALSE, header = TRUE)
rc <- rc[, rownames(meta)]
```

## Preprocessing

```{r}
library(edgeR)

dgelist <- DGEList(
  counts = rc,
  group = meta$cancer_type,
  remove.zeros = TRUE,
  samples = meta,
  genes = rownames(rc)
)

design <- model.matrix(~ 0 + cancer_type, data=meta)
keep <- filterByExpr(dgelist, design=design)
dgelist <- dgelist[keep,,keep.lib.sizes=FALSE]
dgelist <- calcNormFactors(dgelist, method="TMM")
dgelist <- estimateDisp(dgelist, design)

plotBCV(dgelist)
```

## Visualize batch effect

```{r}
library(RColorBrewer)

cols <- brewer.pal(n = length(unique(meta$batch)), name = "Dark2")
names(cols) <- unique(meta$batch)
sample_colors <- cols[meta$batch]

labels <- substr(meta$batch, 6, 7)
plotMDS(dgelist,
        labels = labels,
        col=sample_colors)

```
The second dimension separate batch 5 and 6 from the other batches. Therefore, we should try to perform batch correction in order to remove the batch effects.

## Batch correction

We'll use the `removeBatchEffect` function from `limma` package. Ideally, the input to this function should be close to normal distribution. Therefore, we'll use the `voom` function to normalize the values. Voom fit the distribution of the raw counts and map them to the closest quantile values of a normal distribution.

```{r}
ge <- cpm(dgelist, log=TRUE)
ge <- removeBatchEffect(ge, batch=meta$batch, design=design)

plotMDS(ge,
        labels = labels,
        col=sample_colors)
```

The MDS plot looks better now. We now convert the Ensembl gene IDs to gene symbol and write the file.

```{r}
library(httr)

GENE_SYM_ID_LINK_BASE <- "https://www.genenames.org/cgi-bin/download/custom?col=gd_app_sym&col=gd_pub_eg_id&col=gd_pub_ensembl_id&status=Approved&hgnc_dbtag=on&order_by=gd_app_sym_sort&format=text&submit=submit"
response <- GET(GENE_SYM_ID_LINK_BASE)
gene_sym_id <- data.frame(fread(text=httr::content(response, "parsed"), header=TRUE))
colnames(gene_sym_id) <- c("gene_symbol", "gene_id", "ensg")

gene_sym_id <- gene_sym_id[apply(gene_sym_id == "", 1, sum) == 0,]
gene_sym_id <- gene_sym_id[apply(is.na(gene_sym_id), 1, sum) == 0,]
gene_sym_id <- gene_sym_id[!duplicated(gene_sym_id$ensg), ]
```

```{r}
rownames(gene_sym_id) <- gene_sym_id$ensg
common_genes <- intersect(gene_sym_id$ensg, rownames(ge))
ge <- ge[common_genes,]
rownames(ge) <- gene_sym_id[rownames(ge), "gene_symbol"]
```


```{r}
write.table(round(ge, 4), file="tep_gse68086_voom.txt",
            sep="\t", col.names = NA, quote = FALSE)
```


